Early Stage Assessment

### 2018-12-05 papers/esa/esa.tex
Further edits

The goal of my research is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model.

@@

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{DLMFSphericalandSpheroidalHarmonics, DLMFFerrers, DLMFAssociatedLegendre}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$. While this is equivalent to standard techniques in the litrature concerning spherical harmonics, we approach this with our goal in mind of developing sparse spectral methods on spherical triangles. Thus, we try to use language and formulations that will translate as we move through our steps towards that goal.

\citep{dunkl2014orthogonal} \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is} provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order). This provides an optimal complexity method for evaluating spherical harmonic expansions. This is an alternative to using the explicit expressions that may have benefits, though further investigation is required. \bstodoinline{Reword above}

@@

For each \(l \in \No\) let \(\Dlt\) be any matrix that is a left inverse of \(A_l\), i.e. such that \(\Dlt A_l = I_{2l+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dlt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigP_{-1} := 0 \\
\bigP_{0} := \alpha^0_0 \\
\bigP_{l+1} = -\Dlt (B_l-G_l) \Pl - \Dlt C_l  \,\bigP_{l-1}, \quad l \in \N.
\end{cases}
\end{align}

@@

Then using the operators on the coefficient vectors, the timestepping method is: 
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
where $F, K, G, D$ are all operators corresponding to multiplying by the Coriolis parameter, outward normal vector cross product, gradient and divergence operations respectively. It is important to note that these can be easily constructed due to the relationships between the vector spherical harmonics and scalar spherical harmonics we have outlined, and have \bstodo{Banded-Block-Banded?} a sparse Block-Banded structure.

@@

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the Lanczos algorithm as implemented in the ApproxFun package \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to:\sotodo{Cite Huybrechs reference for connection in literature}
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

By using a similar process to \cite{dunkl2014orthogonal} (p??) \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is} we can construct 2D orthogonal polynomials on the half disk from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\). We denote the half disk by \(\Omega\) where
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad 0 < x < 1, \: -(1-x^2)^{\half} < y < (1-x^2)^{\half}\}.
\end{align}

@@

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

@@

\sotodoinline{Expand bibliography: cite spherical harmonics books, Huybrechs, ECMWF paper, etc.}

>>>>

The goal of my research is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model \cite{wedi2013fast}. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model.

@@

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{DLMFSphericalandSpheroidalHarmonics, DLMFFerrers, DLMFAssociatedLegendre, efthimiou2014spherical, dai2013approximation}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$. While this is equivalent to standard techniques in the litrature concerning spherical harmonics, we approach this with our goal in mind of developing sparse spectral methods on spherical triangles. Thus, we try to use language and formulations that will translate as we move through our steps towards that goal.

\citep{dunkl2014orthogonal} (p75-85) provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order). This provides an optimal complexity method for evaluating spherical harmonic expansions; an alternative to using the explicit expressions that may have benefits, though further investigation is required.

@@

For each \(l \in \No\) let \(\Dlt\) be any matrix that is a left inverse of \(A_l\), i.e. such that \(\Dlt A_l = I_{2l+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dlt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} (p78), which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigP_{-1} := 0 \\
\bigP_{0} := \alpha^0_0 \\
\bigP_{l+1} = -\Dlt (B_l-G_l) \Pl - \Dlt C_l  \,\bigP_{l-1}, \quad l \in \N.
\end{cases}
\end{align}

@@

Then using the operators on the coefficient vectors, the timestepping method is: 
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
where $F, K, G, D$ are all operators corresponding to multiplying by the Coriolis parameter, outward normal vector cross product, gradient and divergence operations respectively. It is important to note that these can be easily constructed due to the relationships between the vector spherical harmonics and scalar spherical harmonics we have outlined, and have a sparse Block-Banded structure.

@@

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the Lanczos algorithm as implemented in the ApproxFun package \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to \cite{huybrechs2010fourier}:
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

By using a similar process to \cite{dunkl2014orthogonal} (p55-56) we can construct 2D orthogonal polynomials on the half disk from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\). We denote the half disk by \(\Omega\) where
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad 0 < x < 1, \: -(1-x^2)^{\half} < y < (1-x^2)^{\half}\}.
\end{align}

@@

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} (p78), which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

@@
